---
title: "Week5a"
author: "Zihao Yu"
date: "2026-02-26"
output:
  html_document: default
  pdf_document: default
---

# 1.How will I tackle the problem?
I will input the data into a spreadsheet and export it as a CSV file. Then, I will read the data into R and perform data cleaning. Finally, I will complete the assignment requirements (comparing user spreadsheets, apply Simpson's Paradox, etc.).

# 2.What data challenges do I anticipate?
During computation and aggregation, prevent incorrect values or improper handling of missing data from affecting the final interpretation. I will verify the understanding of LLM-generated code before applying it.

source: "https://raw.githubusercontent.com/XxY-coder/data607-week5a.Y/refs/heads/main/airlines.csv"


# 3. Transformed data from wide to long format.

```{r}
library(tidyverse)
library(janitor)

airlines_df <- read.csv("https://raw.githubusercontent.com/XxY-coder/data607-week5a.Y/refs/heads/main/airlines.csv")
names(airlines_df)

head(airlines_df)
```

rename and clean the NA values.
```{r}
flights_df <-
  airlines_df %>%
  rename(
    airline = X,
    flight_condition = X.1,
) %>%
  mutate(airline = na_if(airline, "")) %>%
  fill(airline, 
       .direction = "down") %>%
  pivot_longer(
    cols = c("LOS.Angeles", "Phoenix", "San.Diego", "San.Francisco", "Seattle"), 
    names_to = "City", 
    values_to = "Count",
    values_drop_na = TRUE
  )
flights_df
```

# 3.Compared percentage of either delays or arrival rates for two airlines overall

```{r}
overall_delay_compare <- 
  flights_df %>%
  group_by(airline, flight_condition) %>%
  summarise(
    total_flight = sum(Count),
    .groups = 'drop') %>%
  group_by(airline) %>%
  mutate(percentage = round(total_flight / sum(total_flight) * 100,2)) %>%
  filter(flight_condition == "delayed")
overall_delay_compare
```

```{r}
overall_onTime_compare <- 
  flights_df %>%
  group_by(airline, flight_condition) %>%
  summarise(
    total_flight = sum(Count),
    .groups = 'drop') %>%
  group_by(airline) %>%
  mutate(percentage = round(total_flight / sum(total_flight) * 100, 2)) %>%
  filter(flight_condition == "on time")
overall_onTime_compare
```

```{r}
overall_compare <-
  bind_rows(
    overall_onTime_compare,
    overall_delay_compare
  )
overall_compare
```
```{r}
ggplot(
  overall_compare,
  aes(x=airline, y = percentage, fill = flight_condition)
) +
  geom_col(position = "dodge") +
  geom_text(aes(label = percentage),vjust = -0.8, size = 4) +
  labs(title = "Overall Flight Conditions Percentage by Airline", 
       x = "Airline", 
       y = "Percentage (%)") +
  theme_minimal() +
  scale_y_continuous(expand = expansion(mult = c(0, 0.3)))
```

By overall proportion, AMWEST's on-time rate is approximately 89.11%, with a delay rate of about 10.89%; ALASKA's on-time rate is approximately 86.73%, with a delay rate of about 13.27%.

AMWEST's delay rate is approximately 2.38 percentage lower than ALASKA's, while its on-time rate is correspondingly about 2.38 percentage higher. Therefore, MWEST's performance is superior to ALASKA's.

--------------------------------------------------------------------------------
# 4.Compare by each cites.

```{r}
city_delay <-
  flights_df %>%
  group_by(airline, City, flight_condition) %>%
  summarise(city_flights = sum(Count), 
            .groups = 'drop') %>%
  group_by(airline, City) %>%
  mutate(percentage = round(city_flights / sum(city_flights) * 100,2)) %>%
  filter(flight_condition == "delayed")
city_delay
```

```{r}
city_onTime <-
  flights_df %>%
  group_by(airline, City, flight_condition) %>%
  summarise(city_flights = sum(Count), 
            .groups = 'drop') %>%
  group_by(airline, City) %>%
  mutate(percentage = round(city_flights / sum(city_flights) * 100,2)) %>%
  filter(flight_condition == "on time")
city_onTime
```

```{r}
city_compare <-
  bind_rows(
    city_onTime,
    city_delay
  )
city_compare
```

```{r}
ggplot(city_compare, aes(x = City, y = percentage, fill = flight_condition)) +
  geom_col() +
  facet_wrap(~ airline) +
  geom_text(aes(label = round(percentage, 2)),
            position = position_stack(vjust = 0.5),
            size = 4) +
  labs(
    title = "On-time vs Delayed Percentage Across Cities",
    x = "City",
    y = "Percentage (%)",
    fill = "Status"
  ) +
  theme_minimal()
```

According to the city delay rate comparison chart, ALASKA's delay rate is generally lower than AMWEST's. ALASKA's delay rates across five cities range from 5.15% to 16.86%; AMWEST exhibits higher overall delays, with San Francisco at approximately 28.73% and Seattle at 23.28%. Comparing city-specific delay rates, ALASKA maintains lower delays.

# 5.Conclusion
When comparing by city, ALASKA had lower delay rates than AMWEST in all five cities; however, when combining all cities, AMWEST's overall delay rate was lower. This is a typical example of Simpson's paradox.

